### 2023-01-22 chapter6/chapter.tex
.
post-reviews, pre supervisor feedback

The performance of the discontinuous Galerkin scheme has been well-studied in some contexts (\cite{c99}, \cite{nodalDGBook}~Section~5.5), and it is well-known that the discontinuous Galerkin scheme performs remarkably well on problems with smooth solutions. Here, we mostly focus on investigating the numerical performance of the schemes on problems with non-smooth solutions. The purpose is to emphasise the probabilistic interpretation and positivity-preserving properties of the QBD-RAP scheme. In the stochastic modelling community it is common to have problems with discontinuities, such as non-smooth initial conditions, hence the type of problems we investigate in this chapter are of relevance. As we will see (in Section~\ref{sec: transient approx}), even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities or non-smooth behaviour into the problem. A very specific set of conditions on the initial density and point masses must hold for the transient distribution to remain continuous as it evolves over time \citep{bo2014,bop2020}. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues (see, for example, Section~\ref{sec: ffq num}). Limiting distributions of fluid queues are smooth, however.

@@

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance which allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which means that the solution is known, and we do not need to resort to simulation as a \emph{ground-truth}. Furthermore, the travelling wave problems allow us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates approximations to the limiting distribution of the queue, Section~\ref{sec: transient approx} investigates approximations to the transient distribution of the queue for two initial conditions, and Section~\ref{sec: hit approx} investigates approximations the first hitting time of the fluid level on the set \(\{0,1\}\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num}, compute approximations to \(\mathbb \Psi\) and ultimately, approximate the distribution of \(\{(X(\zeta_{Y}(\{0\})),\varphi(\zeta_{Y}(\{0\})))\}\) at time \(\zeta_{Y}(\{0\})\) at which \(\{Y(t)\}\) first returns to \(0\).

@@

% The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance. This allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics of the problem. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which allows us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates the ability of the schemes to approximate the limiting distribution of the model, Section~\ref{sec: transient approx} investigates the ability of the schemes to approximate the transient distribution of the model for two initial conditions, and Section~\ref{sec: return approx} investigates the ability of the schemes to approximate the first hitting time of the fluid level on the boundary of the interval \([0,1]\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num} and approximate the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{Y(t)\}\) first returns to \(0\).

@@

\begin{example}\label{ex: another example}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) of Example~\ref{ex: another example} for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves of the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}%To estimate the asymptotic rate of convergence, we fit a linear model of \(\log_{10}(error) = \beta_0 + \beta_1\log_{10}(Dim)\) to the the last 6 data points on the curves above, which corresponds to a rate of convergence \(error \propto N^{-\beta_1}\). The estimate of \(\beta_1\) was approximately -1.0 in all cases.}
		\label{fig: fun 2 ks error qbdrap closing vecs}
	\end{figure} 
	%\exampleFloatBarrier
	First consider approximating the initial condition with density function \(2\times 1(x<0.5)\), where \(1(\cdot)\) is the indicator function. In the left-hand side of Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} we plot the Kolmogorov-Smirnov error between the true CDF and the CDF constructed via the QBD-RAP approximation with the three different closing vectors. Interestingly, in this case, the unnormalised closing operator outperforms the other two at low dimensions. In this case, it just so happens that the error due to truncation of the unnormalised reconstruction constructively interferes with other errors to cause the error to be lower at low dimensions for the unnormalised scheme.

@@

\begin{example}\label{ex: al}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) of Example~\ref{ex: al} for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves of the error curves of the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 4 ks error qbdrap closing vecs}
	\end{figure}
	Now consider approximating the initial density \(f(x)=1\) for \(x\in[0,1)\). Observing Figure~\ref{fig: fun 4 ks error qbdrap closing vecs} of the KS error and \(L^2\) error of the PDF we now see that the normalised reconstructions outperform the unnormalised reconstruction. This suggests that, in this case, the `folding' of the closing operator about \(\Delta\) has greatly increased the ability of the reconstruction to approximate this initial distribution.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/pdfs_formatted.pdf}
	\caption{Approximations to the PDF \(f(x)=1,\, x\in[0,1)\) (orange dotted line) for Example~\ref{ex: al}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (blue solid line) and normalised closing operator (green dash-dotted line).}
	\label{fig: pdf reconstructed}
\end{figure} 
%\exampleFloatBarrier
Figure~\ref{fig: pdf reconstructed} also shows that both closing operators fail to approximate the initial condition well at the right-hand side of the interval. A potential area for further research is to determine if there is a different reconstruction method or another closing operator which could alleviate this issue. 
\end{example}

\begin{example}\label{ex: aasmm}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error for the PDFs (right) of Example~\ref{ex: aasmm}, for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves for the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 6 ks error qbdrap closing vecs}
	\end{figure}
	Next we consider approximating the initial distribution with density \(-6x^2+6x\) on \([0,1)\). Observing the left-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the KS error against the dimension of the matrix exponential approximation for the three closing operators, we once again see that the unnormalised closing operator performs the worst, while the performance of the two normalised closing operators is indistinguishable. However, if we instead look at the right-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the \(L^2\) error between the reconstructed PDF and the true PDF, then the unnormalised closing operator performs better than the two normalised ones for dimension 5 and above.

	\begin{figure}[h]
		\centering
		\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/pdfs_formatted.pdf}
		\caption{Approximations to the PDF \(f(x)=-6x^2+6x\) (orange dotted line) for Example~\ref{ex: aasmm}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (blue solid line) and normalised closing operator (green dash-dotted line).} 
		\label{fig: pdf reconstructed quadratic}
	\end{figure} 
	%\exampleFloatBarrier
The fact that the unnormalised closing operator outperforms the two normalised ones can be explained by the `folding' of the normalised operators around \(\Delta\). In Figure~\ref{fig: pdf reconstructed quadratic} we plot the reconstructions obtained using the unnormalised and normalised closing operators, along with the true PDF, \(-6x^2+6x\). In Figure~\ref{fig: pdf reconstructed quadratic} at the left of the plots, we observe that the normalised closing operator over-estimate the density function, whereas the unnormalised closing operator looks to be doing better. The `folding' in the normalised closing operator manifests itself as more mass at the left of these plots compared to the unnormalised closing operator. 
\end{example}

\begin{example}\label{ex: kjfk}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) for Example~\ref{ex: kjfk} for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves for the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 7 ks error qbdrap closing vecs}
	\end{figure}
	%\exampleFloatBarrier
	Lastly, we consider approximating the initial distribution with PDF \(3e^{-3x}/(1-e^{-3})\). This density function is at a maximum at the left of the region. Considering what we have learnt so far about the unnormalised operator underestimating in this region, we expect that the unnormalised closing operator will perform relatively poorly in this case. Indeed, observing Figures~\ref{fig: fun 7 ks error qbdrap closing vecs} we see that this is the case; the normalised closing operators perform relatively well compared to the unnormalised closing operator, as measured by both error metrics (KS statistic and \(L^2\) norm on the PDF). Plots of the PDFs (omitted), confirm that this is due to the loss of mass at the left-hand side of the region due to the truncation. 
\end{example}

@@

\begin{example}\label{ex: wave 4}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.75cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/wave/fun4/L1_cell_probs.pdf}
		\caption{Cell-wise error defined in (\ref{eqn: cell errors}) for the travelling wave in Example~\ref{ex: wave 4}. Plotted are the cell-wise errors for the DG (blue solid line), DG-lim scheme (green dashed line), uniformisation (orange dashed line), QBD-RAP (purple dotted line) and DG-lin-lim scheme (gold dashed line) schemes, versus the dimension of the approximation. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 4 wave cp} 
	\end{figure}
First consider the initial condition with PDF \(1(x<1)\). The level of the fluid queue is uniformly distributed over the first cell. For the DG-based schemes (including the uniformisation scheme) the initial condition can be represented exactly whereas, for the QBD-RAP scheme, it cannot. Thus, in this case, there is no discretisation error in constructing the initial condition for the DG and uniformisation schemes. Moreover, at time \(t=4\), the solution is \(f(x,4)=1(x\in[4,5))\). The projections related to the DG-based schemes can represent this solution exactly too, hence we might expect these schemes to work well here.

@@

\begin{example}\label{ex: wave 6}
	Now consider an initial distribution which is truncated Gaussian with mean 2.5 and standard deviation 0.5, and truncated at the boundaries, 0 and 10;
	\begin{align}
		\mu([0,x)) := \mathbb P(X(0)\leq x, \varphi(0)=1)=\cfrac{\Phi((x-2.5)/0.5)-\Phi(-2.5/0.5)}{\Phi(7.5/0.5)-\Phi(-2.5/0.5)}, \label{eqn: kdjfksdf}
	\end{align}
	for \(x\in[0,10]\) and where \(\Phi(x)\) is the CDF of the standard normal distribution.

>>>>

The performance of the discontinuous Galerkin scheme has been well-studied in some contexts (\cite{c99}, \cite{nodalDGBook}~Section~5.5), and it is well-known that the discontinuous Galerkin scheme performs remarkably well on problems with smooth solutions. Here, we mostly focus on investigating the numerical performance of the schemes on problems with non-smooth solutions. The purpose is to emphasise the probabilistic interpretation and positivity-preserving properties of the QBD-RAP scheme. In the stochastic modelling community it is common to have problems with discontinuities, such as non-smooth initial conditions, hence the type of problems we investigate in this chapter are of relevance. As we will see (in Section~\ref{sec: transient approx}), even if a fluid-queue is initialised with a smooth initial density, the boundary dynamics may induce transient discontinuities or non-smooth behaviour into the problem. A very specific set of conditions on the initial density and point masses must hold for the transient distribution to remain continuous as it evolves over time \citep{bo2014,boz2022}. Further, it is possible that discontinuities are present in the performance measures of fluid-fluid queues (see, for example, Section~\ref{sec: ffq num}). Limiting distributions of fluid queues are smooth, however.

@@

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance which allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which means that the solution is known, and we do not need to resort to simulation as a \emph{ground-truth}. Furthermore, the travelling wave problems allow us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates approximations to the limiting distribution of the queue, Section~\ref{sec: transient approx} investigates approximations to the transient distribution of the queue for two initial conditions, and Section~\ref{sec: hit approx} investigates approximations the first hitting time of the fluid level on the set \(\{0,1\}\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num}, compute approximations to \(\mathbb \Psi\) and ultimately, approximate the distribution of \(\{(X(\zeta_{W}(\{0\})),\varphi(\zeta_{W}(\{0\})))\}\) at time \(\zeta_{W}(\{0\})\) at which \(\{\dot W(t)\}\) first returns to \(0\).

\paragraph{On the application of the QBD-RAP approximation to approximating operators of fluid-fluid queues.}
A few remarks about applying the QBD-RAP approximation to approximate fluid queues and operators of fluid-fluid queues from \cite{bo2014} are in order.

Despite the construction of the QBD-RAP in Chapter~\ref{sec: construction and modelling} and the proof of convergence to the fluid queue spanning Chapters~\ref{sec: conv} and \ref{ch: global results} being relatively technical, the actual computations involved are relatively simple once we have the generator of the QBD-RAP. Construction of the generator matrix of the QBD-RAP, \(\bs B\), is a matter of obtaining the parameters \(\bs\alpha,\,\bs S\) and \(\bs s\) of a concentrated matrix exponential distribution from \cite{hht2020}, computing the matrix \(\bs D\) as in the last part of Section~\ref{sec:kkkrjrjrjjrj}, then assembling the parameters into the tridiagonal form of the generator as in (\ref{eqn: BBBB})-(\ref{eqn: BBBB3}) and including boundary conditions as in (\ref{eqn: BBBBbndry})-(\ref{eqn: BBBBbndry2}). The initial vector of the QBD-RAP, \(\bs a=(\bs a_{k,i})_{i\in\calS,k\in\mathcal K}\), can be constructed as discussed in Section~\ref{sec: initial conditions}, perhaps using numerical integration depending on the choice of initial condition. After constructing the generator and initial condition, approximations to the fluid queue can be obtained relatively easily. For example, the expected orbit positions in each phase and level, \(\mathbb E[\bs A(t)1(L(t)=\ell, \phi(t)=i)]\), are found by computing 
\[(\mathbb E[\bs A(t)1(L(t)=\ell, \phi(t)=i)])_{i\in\calS, \ell \in\mathcal K}=\bs a \exp(\bs Bt),\]
which, in this chapter, is computed numerically via the SSPRK integration scheme in Section~\ref{sec: time integration}. The approximation to the transient density is obtained from \(\mathbb E[\bs A(t)1(L(t)=\ell, \phi(t)=i)]\) as discussed in Section~\ref{sec: closing}.

To approximate operators of fluid-fluid queues with the QBD-RAP  we need to approximate the operator \(\mathbb R\), defined in Lemma~\ref{lemma: D(s)}. Recall from Section~\ref{sec: approx r}, in the context of the DG scheme we could approximate \(\mathbb R\) as a projection. In the context of the QBD-RAP scheme it is less clear how one can approximate \(\mathbb R\), in general. If the rate functions \(r_i(x)\) are piecewise constant with \(r_i(x)=r_i^k\) on cell \(\calD_{k,i}\), then a suitable approximation for \(\mathbb R^k_i\) is \(1/|r_i^k|\bs I_{p}\), a diagonal matrix with \(1/|r_i^k|\) down the diagonal -- this is the case in the numerical experiments in Chapter~\ref{sec: numerics}. For more general cases we could consider approximating \(r_i(x)\) by a piecewise constant function, then apply the method above.

Next, the approximation \(\bs \Psi\) is computed by solving the matrix Riccati equation 
\begin{align}\label{eqn:RiccatiPsi2222aa}
    \bs D^{+-}
+ \bs \Psi   \bs D^{-+}\bs \Psi
+   \bs D^{++}\bs \Psi
+ \bs \Psi  \bs D^{--}
= \bs 0,
\end{align}
where 
\[\bs D^{mn} = \bs{R}^m (\bs{B}^{mn} + \bs{B}^{m0}(\bs B^{00})^{-1}\bs{B}^{0n}), \, m,n\in\{+,-\},\]
and \(\bs{B}^{mn},\, m,n\in\{+,-,0\}\) are matrices of parameters extracted from the QBD-RAP which approximate \(\mathbb B^{mn}\). For example, \(\bs B^{+-}\) is obtained by extracting all the elements of \(\bs B\) which correspond to starting in phase \(i\) in level \(k\) and ending in phase \(j\) in level \(\ell\) and have \(r_i(x)>0\) for \(x\in\calD_{k,i}\) and \(r_j(x)<0\) for \(x\in\calD_{\ell,j}\). Given the stochastic interpretation of the QBD-RAP scheme, and assuming the rate functions \(r_i(x)\) are piecewise constant with \(r_i(x)=r_i^k\) on each cell \(\calD_{k,i}\), the QBD-RAP approximation of the fluid-fluid queue forms a RAP-modulated fluid process \citep{p2019,bgnp2021}. Hence, the approximation \(\bs \Psi\) has a stochastic interpretation in terms of the first-return operator of the RAP-modulated fluid process. This is important. Equation (\ref{eqn:RiccatiPsi2222aa}) determines, but does not define \(\bs \Psi\). When we use the DG scheme to approximate \(\mathbb \Psi\) we still solve (\ref{eqn:RiccatiPsi2222aa}) (with the DG approximation to the generator \(\mathbb B\)), however, this does not define what the solution means in terms of the fluid-fluid queue (other than it is the solution to (\ref{eqn:RiccatiPsi2222aa})). In comparison, there is a rigorous definition of \(\bs \Psi\) when the QBD-RAP scheme is used in the approximation \citep{p2019,bgnp2021}.

By Proposition~4.3 of \cite{p2019}, for the QBD-RAP scheme the approximation \(\bs \Psi\) has the property \(\bs \Psi \bs 1=\bs 1\) as long as the event that level process of the RAP-modulated fluid returns to \(0\) in finite time occurs with probability 1.

Third, note that unlike the positivity preserving DG schemes which use slope limiting which are not linear operators in the presence of oscillations, the operator obtained via the QBD-RAP approximation scheme is linear.

{Lastly, a comment on expected accuracy of the QBD-RAP scheme.} For \(p\) odd, \cite{hht2020} estimate that the variance of the class of CMEs decreases at order \(1/p^{2.14}\) asymptotically. Hence, we might expect the rate of convergence of the QBD-RAP scheme to be approximately order \(1/p^2\) for certain problems. This is in line with the positivity preserving DG scheme mentioned in~\ref{sec: anoth pos pres}. %The justification for considering representations of odd orders only is that the variance of CME representations of orders \(2p\) and \(2p-1\) are relatively similar and therefore have similar abilities to represent the delta function \citep{hht2020}. Hence, if we construct a QBD-RAP approximation with a representation of order \(2p\) we expect it to perform only marginally better than an approximation constructed with representations of order \(2p-1\). However, the computational cost of the latter is lower, so we opt for the order \(2p-1\) representation.

@@

% The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance. This allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics of the problem. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which allows us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates the ability of the schemes to approximate the limiting distribution of the model, Section~\ref{sec: transient approx} investigates the ability of the schemes to approximate the transient distribution of the model for two initial conditions, and Section~\ref{sec: return approx} investigates the ability of the schemes to approximate the first hitting time of the fluid level on the boundary of the interval \([0,1]\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num} and approximate the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{W(t)\}\) first returns to \(0\).

@@

\begin{example}\label{ex: another example}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun2/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) of Example~\ref{ex: another example} for the three closing vectors considered; unnormalised (green solid line), non-linear normalised (blue dashed line) and normalised (orange dash-dotted line). The error curves of the non-linear normalised and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}%To estimate the asymptotic rate of convergence, we fit a linear model of \(\log_{10}(error) = \beta_0 + \beta_1\log_{10}(Dim)\) to the the last 6 data points on the curves above, which corresponds to a rate of convergence \(error \propto N^{-\beta_1}\). The estimate of \(\beta_1\) was approximately -1.0 in all cases.}
		\label{fig: fun 2 ks error qbdrap closing vecs}
	\end{figure} 
	%\exampleFloatBarrier
	First consider approximating the initial condition with density function \(2\times 1(x<0.5)\), where \(1(\cdot)\) is the indicator function. In the left-hand side of Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} we plot the Kolmogorov-Smirnov error between the true CDF and the CDF constructed via the QBD-RAP approximation with the three different closing vectors. Interestingly, in this case, the unnormalised closing operator outperforms the other two at low dimensions. In this case, it just so happens that the error due to truncation of the unnormalised reconstruction constructively interferes with other errors to cause the error to be lower at low dimensions for the unnormalised scheme.

@@

\begin{example}\label{ex: al}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) of Example~\ref{ex: al} for the three closing vectors considered; unnormalised (green solid line), non-linear normalised (blue dashed line) and normalised (orange dash-dotted line). The error curves of the error curves of the non-linear normalised and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 4 ks error qbdrap closing vecs}
	\end{figure}
	Now consider approximating the initial density \(f(x)=1\) for \(x\in[0,1)\). Observing Figure~\ref{fig: fun 4 ks error qbdrap closing vecs} of the KS error and \(L^2\) error of the PDF we now see that the normalised reconstructions outperform the unnormalised reconstruction. This suggests that, in this case, the `folding' of the closing operator about \(\Delta\) has greatly increased the ability of the reconstruction to approximate this initial distribution.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/pdfs_formatted.pdf}
	\caption{Approximations to the PDF \(f(x)=1,\, x\in[0,1)\) (blue dashed line) for Example~\ref{ex: al}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (green solid line) and normalised closing operator (orange dash-dotted line).}
	\label{fig: pdf reconstructed}
\end{figure} 
%\exampleFloatBarrier
Figure~\ref{fig: pdf reconstructed} also shows that both closing operators fail to approximate the initial condition well at the right-hand side of the interval. A potential area for further research is to determine if there is a different reconstruction method or another closing operator which could alleviate this issue. 
\end{example}

\begin{example}\label{ex: aasmm}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error for the PDFs (right) of Example~\ref{ex: aasmm}, for the three closing vectors considered; unnormalised (green solid line), non-linear normalised (blue dashed line) and normalised (orange dash-dotted line). The error curves for the non-linear normalised and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 6 ks error qbdrap closing vecs}
	\end{figure}
	Next we consider approximating the initial distribution with density \(-6x^2+6x\) on \([0,1)\). Observing the left-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the KS error against the dimension of the matrix exponential approximation for the three closing operators, we once again see that the unnormalised closing operator performs the worst, while the performance of the two normalised closing operators is indistinguishable. However, if we instead look at the right-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the \(L^2\) error between the reconstructed PDF and the true PDF, then the unnormalised closing operator performs better than the two normalised ones for dimension 5 and above.

	\begin{figure}[h]
		\centering
		\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/pdfs_formatted.pdf}
		\caption{Approximations to the PDF \(f(x)=-6x^2+6x\) (blue dashed line) for Example~\ref{ex: aasmm}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (green solid line) and normalised closing operator (orange dash-dotted line).} 
		\label{fig: pdf reconstructed quadratic}
	\end{figure} 
	%\exampleFloatBarrier
The fact that the unnormalised closing operator outperforms the two normalised ones can be explained by the `folding' of the normalised operators around \(\Delta\). In Figure~\ref{fig: pdf reconstructed quadratic} we plot the reconstructions obtained using the unnormalised and normalised closing operators, along with the true PDF, \(-6x^2+6x\). In Figure~\ref{fig: pdf reconstructed quadratic} at the left of the plots, we observe that the normalised closing operator over-estimate the density function, whereas the unnormalised closing operator looks to be doing better. The `folding' in the normalised closing operator manifests itself as more mass at the left of these plots compared to the unnormalised closing operator. 
\end{example}

\begin{example}\label{ex: kjfk}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) for Example~\ref{ex: kjfk} for the three closing vectors considered; unnormalised (green solid line), non-linear normalised (blue dashed line) and normalised (orange dash-dotted line). The error curves for the non-linear normalised and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 7 ks error qbdrap closing vecs}
	\end{figure}
	%\exampleFloatBarrier
	Lastly, we consider approximating the initial distribution with PDF \(3e^{-3x}/(1-e^{-3})\). This density function is at a maximum at the left of the region. Considering what we have learnt so far about the unnormalised operator underestimating in this region, we expect that the unnormalised closing operator will perform relatively poorly in this case. Indeed, observing Figures~\ref{fig: fun 7 ks error qbdrap closing vecs} we see that this is the case; the normalised closing operators perform relatively well compared to the unnormalised closing operator, as measured by both error metrics (KS statistic and \(L^2\) norm on the PDF). Plots of the PDFs (omitted), confirm that this is due to the loss of mass at the left-hand side of the region due to the truncation. 
\end{example}

@@

\begin{example}\label{ex: wave 4}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.75cm 0.8cm 0.25cm 1.25cm},clip]{chapter6/figs/wave/fun4/L1_cell_probs.pdf}
		\caption{Cell-wise error defined in (\ref{eqn: cell errors}) for the travelling wave in Example~\ref{ex: wave 4}. Plotted are the cell-wise errors for the DG (blue solid line), DG-lim scheme (green dashed line), uniformisation (orange dashed line), QBD-RAP (purple dotted line) and DG-lin-lim scheme (gold dashed line) schemes, versus the dimension of the approximation. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 4 wave cp} 
	\end{figure}
First consider the initial condition with PDF \(1(x<1)\), a uniform distribution over \((0,1)\). The level of the fluid queue is uniformly distributed over the first cell. For the DG-based schemes (including the uniformisation scheme) the initial condition can be represented exactly whereas, for the QBD-RAP scheme, it cannot. Thus, in this case, there is no discretisation error in constructing the initial condition for the DG and uniformisation schemes. Moreover, at time \(t=4\), the solution is \(f(x,4)=1(x\in[4,5))\). The projections related to the DG-based schemes can represent this solution exactly too, hence we might expect these schemes to work well here.

@@

\begin{example}\label{ex: wave 6}
	Now consider an initial distribution which is truncated Gaussian with mean 2.5 and standard deviation 0.5, and truncated at the boundaries, 0 and 10;
	\begin{align}
		\mu([0,x)) = \mathbb P(X(0)\leq x, \varphi(0)=1)=\cfrac{\Phi((x-2.5)/0.5)-\Phi(-2.5/0.5)}{\Phi(7.5/0.5)-\Phi(-2.5/0.5)}, \label{eqn: kdjfksdf}
	\end{align}
	for \(x\in[0,10]\) and where \(\Phi(x)\) is the CDF of the standard normal distribution.
